---
title: "TidyTuesday, week 30"
author: "Marco Lombardi"
date: "`r Sys.Date()`"
output: 
  html_document:
    number_sections: no
    self_contained: no
    theme: yeti
    toc: no
---    

\ 

# US Droughts Data

Dry and wet cycles are dependent on a number of factors including geography, topology and climate drivers. Drought phases in the US can be explained by changes in the track of extratropical cyclones, which can occur during climate cycles such as the El Niño-Southern Oscillation, or ENSO, as well as the North Atlantic Oscillation, Pacific Decadal Oscillation, and Atlantic multidecadal oscillation. The country's contiguous western and especially southwestern region has experienced widespread drought since about year 2000 (https://en.wikipedia.org/wiki/Droughts_in_the_United_States).


\ 

**TidyTuesday Link**

https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-07-20/readme.md

\ 

**Data**

2001-2021 US droughts data are collected by the [US Drought Monitor](https://droughtmonitor.unl.edu/DmData/DataDownload.aspx). 

> The U.S. Drought Monitor is jointly produced by the National Drought Mitigation Center at the University of Nebraska-Lincoln, the United States Department of Agriculture, and the National Oceanic and Atmospheric Administration. Map courtesy of NDMC.


```{r settings, echo=F}
knitr::opts_chunk$set(results='markdown', 
                      message = F,
                      echo=F,
                      width=200,
                      fig.width=12, 
                      fig.height=9, 
                      dev.args = list(pointsize=14), 
                      dpi=75,
                      fig.path = "./figures/")

library(tidytuesdayR) 
library(data.table)
library(kableExtra)
library(ggplot2)
library(gganimate)
```

```{r load, include=FALSE}
# Either ISO-8601 date or year/week works!
tuesdata <- tidytuesdayR::tt_load('2021-07-20')
tuesdata <- tidytuesdayR::tt_load(2021, week = 30)
dt_drought <- data.table(tuesdata$drought)

dt_drought<- dt_drought[, drought_desc := factor(drought_lvl, 
                                                 levels = c("None", "D0", "D1", "D2", "D3", "D4"), 
                                                 labels = c("no drought", 
                                                          "abnormally dry", 
                                                          "moderate drought", 
                                                          "severe drought", 
                                                          "extreme drought",
                                                          "exceptional drought")) ]
```

```{r droughts_data}
kable(head(dt_drought, 12),
      format = "html",
      caption= "droughts data") %>%
  kable_styling(bootstrap_options = "striped", position = "center")
```

```{r droughts_levels}
data.table(levels = c("None", "D0", "D1", "D2", "D3", "D4"), 
           labels = c("no drought",
                      "abnormally dry", 
                      "moderate drought", 
                      "severe drought", 
                      "extreme drought",
                      "exceptional drought")) %>%
  kable(format = "html",
        caption= "drought levels") %>%
  kable_styling(bootstrap_options = "striped", position = "center")
```

\ 

```{r aggregation_per_level}
# Droughts occurrences per year and month.
dt_ym<- dt_drought[, .(sum_area_pct = round(sum(area_pct),6)/uniqueN(map_date),
                         sum_pop_pct = round(sum(pop_pct),6)/uniqueN(map_date)),
                        .(map_year= year(valid_start), 
                          map_month= month(valid_start), 
                          state_abb, 
                          drought_lvl,
                          drought_desc)]

# states affected by extreme drought (D3) or exceptional drought (D4).
dt_ym_D3 <- dt_ym[drought_lvl %in% c("D3","D4"), 
                      .(sum_area_pct = sum(sum_area_pct), 
                        sum_pop_pct = sum(sum_pop_pct)),
                      .(map_year, 
                        map_month,
                        state_abb)]

dt_ym_D3 <- dt_ym_D3[, map_ym := paste(
  map_year, 
  ifelse(map_month %in% 1:9, paste0("0", as.character(map_month)), map_month), 
  sep = "-") ]

```

\ 

# Population affected by "extreme drought" or "exceptional drought"

```{r ym_animate}
p <- 
dt_ym_D3[map_year %in% 2001:2021,] %>%
  ggplot(aes(x= sum_area_pct/100, 
             y=sum_pop_pct/100, 
             fill= sum_pop_pct, 
             label=state_abb,
             size = sum_pop_pct/20)) +
  geom_abline(slope = 1, intercept = 0, linetype= 2) + 
  geom_point(show.legend = F) + 
  geom_label(show.legend = F) +
  labs(x= "Area affected", y = "Population affected") + 
  scale_x_continuous(breaks= seq(0,1,.1), labels = scales::label_percent(accuracy = 1)) +
  scale_y_continuous(breaks= seq(0,1,.1), labels = scales::label_percent(accuracy = 1)) + 
  scale_fill_gradient(low = "#007852", high = "#fb0f03") + 
  theme_bw(base_size = 18) + 
  scale_size(range = c(0,15)) 
  
anim <- p + 
  transition_states(map_ym, transition_length = 3, state_length = 0)  + 
  labs(title = 'States affected by extreme drought or exceptional drought \n Year-month: {closest_state}', 
       x= "Area affected",
       y = "Population affected") +
  ease_aes('linear')

```


```{r gif}
animate(anim, nframes = 2655, fps = 15,  width = 1000, height = 562,
        renderer = gifski_renderer("drought_ym.gif"))
```


```{r mp4}
animate(nframes = 2655, fps = 15,  width = 1000, height = 562,
  anim + enter_fade() + exit_fly(y_loc = 1),
  renderer = av_renderer(file = "./drought_ym.mp4", 
                         audio="The Beatles - Rain.mp3")
)
```

